---
title: "EDA.rmd"
author: "Carlo van Buiten"
date: "9/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
```

# Research question
Is cell uniformity (shape and size) a good predictor for the malignancy of breast cancer?

Attributes definitions: https://www.rai-light.com/docs/BCD_User_Manual_v01.pdf


```{r}
data <- read.csv("breast-cancer-wisconsin.data", na.strings="?")
colnames(data) <- c("id", "Clump_Thickness", "Cell_Size_Uniformity", "Cell_Shape_Uniformity", "Marginal_Adhesion",
                    "Single_Epithelial_Cell_Size", "Bare_Nuclei", "Bland_Chromatin", "Normal_Nucleoli", "Mitoses",
                    "Class")

head(data, 10)
summary(data)
```

Let us take a look at the highest correlating features.
```{r}
cor_data <- cor(data, use = "complete.obs")
cor_data <- as.data.frame(cor_data)
cor_cols <- names(cor_data)[which(cor_data$Class > 0.8)]
cor_cols 
```
There are 3 attributes that correlate with malignancy for more than 80%, 2 of which are pertaining to cell uniformity. This bodes well for the hypothesis that cell uniformity could potentially be a good predictor of malignant breast cancer.

```{r}
plot_data <- data
plot_data$Class[plot_data$Class == 2] <- "B"
plot_data$Class[plot_data$Class == 4] <- "M"

ggplot(data=plot_data, mapping = aes(x = Class, y = Cell_Size_Uniformity)) +
  geom_jitter(width = 0.05, height = 0.3, col="red", alpha = 0.3) +
  theme_minimal()
```

```{r}
ggplot(data=plot_data, mapping = aes(x = Class, y = Cell_Shape_Uniformity)) +
  geom_jitter(width = 0.05, height = 0.3, col="green", alpha = 0.3) +
  theme_minimal()
```

```{r}
ggplot(data=plot_data, mapping = aes(x = Class, y = Bare_Nuclei)) +
  geom_jitter(width = 0.05, height = 0.3, col="blue", alpha = 0.3) +
  theme_minimal()
```






